Asymmetric tail-association in extreme climatic events affects population-extinction risk

### 2019-10-12 MainText.Rmd
editing intro...continued

***Shyamolina: species reordering is used to access the change in community 
in response to extreme events or global environmental changes, it relates to change 
in relative abundances of dominant species (mainly). For example, as different species have different 
sensitivity levels to extreme condition (means the extreme threshold is not 
constant for all the species in a community), some years with extreme drought favors say c3 grass 
but in the next year c4 grass could be dominant. Such alteration in community composition has "impact on 
ecosystem function, leading to an extreme ecological response". For detailed overview you may see 
@hoover2014resistance, Jones et al. Ecol. lett. 2017 (doi: 10.1111/ele.12864), 
Smith et al., Ecology 2009 (doi:10.1890/08-1815.1).
Actually LAN K. KNAPP and MELINDA D. SMITH did a lot of work on this to access changes in 
plant community response to extreme events 
(this review article: Felton et al. 2017, Phil. Trans. R. Soc. B 372: 20160142 is a nice summary).
-->
<!--***DAN: Shyamolina, you have obviously done a lot of reading to situate your manuscript
in the appropriate literature, and that is great!-->
through 
life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annual survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) can 
make a population vulnerable to extirpation.
<!--All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction.--> 
<!--***DAN: Shyamolina, demogra[hic stochasticity is a specific thing. I suspect you may
just want to say instead something like "All these studies indicated that environmental stochasticity (through ECEs) can make a population vulnerable to extirpation."] -->

@@

<!--concept for correlation in the extreme tail: tail-association-->
Though earlier literature considered the influence of spatial and/ or temporal 
correlations in environmental variables on extinction risk of a metapopulation, to 
our knowledge the separate influence on extinction risk of "tail associations" between
environmental time series in different locations has hardly been considered. "Tail associations" are
the association between two random variables in their extreme values. If two positively correlated 
variables are more associated in their smaller values than larger, they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). Similarly, if two positively correlated 
variables are more associated in their larger values than smaller, they are said to have stronger
"right-tail" or "upper-tail" association (Figure \ref{fig_intro_cop}C, D). Figure \ref{fig_intro_cop} shows 
two random variables can have a wide range of assciation from extreme to moderate degree in their extreme tails, but
with a fixed Kendall's $\tau$.

” But this is not the topic sentence. The topic sentence comes second and is something like “Tail associations are…” or “We say two positively associated variables are more right- than left-tail associated if…”. Then the para continues, its point being to explain what asymmetric tail associations are. You can use a pedagogical figure. One thing you probably want to include is the logic (see BIVAN) that if extreme events are getting stronger and more extreme/intense events are also more spatially widespread, then it seems reasonable to expect that spatial patterns of tail associations in environmental variables will also be changing. You may also want to mention (one sentence only) that we use the statistical concept of the copula to control tail dependence in our simulations – more will be said about copulas in the Methods.

@@

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
We recently [@Ghosh_copula] used this tail association feature using "copula" - a statistical 
tool broadly used in modeling dependence between variables - to estimate extinction risk of a density-independent 
metapopulation model. Present manuscript is an extension of this concept 
in estimating the extinction risk beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
Following we will reintroduce the concept of copula in a manner that follows @Ghosh_copula.

<!--brief intro for copula: I am not sure whether it is needed in this paper or not?-->
<!--***DAN: Shyamolina, to answer your above Q, well, you use Clayton and survival 
Clayton in your simulations, so you probably have to explain copulas enough to explain those.-->
Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail association. 
Tail association, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) association is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger association in either tail causes asymmetric copula 
structure for the joint distribution of two variables. For example, Clayton and Survival Clayton are 
two copula families 
<!--***DAN: Shyamolina, we cannot just mention these, no one will know what they are.
As I said I think we should intro tail associations without copulas, and then subsequently link 
to copulas afterwards.-->
with asymmetric tail that are used to model exclusively 
left and right tail association between variables, respectively.
From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail association between vital rates when temperature is too cold than normal) and 
thereby inflate the population viability. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}). Though tail-association 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail association have never been considered 
in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

>>>>

***Shyamolina: species reordering is used to access the change in community 
in response to extreme events or global environmental changes, it relates to change 
in relative abundances of dominant species (mainly). For example, as different species have different 
sensitivity levels to extreme condition (means the extreme threshold is not 
constant for all the species in a community), some years with extreme drought favors say c3 grass 
but in the next year c4 grass could be dominant. Such alteration in community composition has "impact on 
ecosystem function, leading to an extreme ecological response". For detailed overview you may see 
@hoover2014resistance, Jones et al. Ecol. lett. 2017 (doi: 10.1111/ele.12864), 
Smith et al., Ecology 2009 (doi:10.1890/08-1815.1).
Actually LAN K. KNAPP and MELINDA D. SMITH did a lot of work on this to access changes in 
plant community response to extreme events 
(this review article: Felton et al. 2017, Phil. Trans. R. Soc. B 372: 20160142 is a nice summary).
-->
<!--***DAN: Shyamolina, you have obviously done a lot of reading to situate your manuscript
in the appropriate literature, and that is great!-->
through life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annual survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) can 
make a population vulnerable to extirpation.
<!--All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction.-->

@@

<!--concept for correlation in the extreme tail: tail-association-->
Though earlier literature considered the influence of spatial and/ or temporal 
correlations in environmental variables on extinction risk of a metapopulation, to 
our knowledge the separate influence on extinction risk of "tail associations" between
environmental time series in different locations has hardly been considered. "Tail associations" are
the association between two random variables in their extreme values. If two positively correlated 
variables are more associated in their smaller values than larger, they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). Similarly, if two positively correlated variables are more associated in their larger values than smaller, they are 
said to have stronger "right-tail" or "upper-tail" association (Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows two random variables can have a wide 
range of assciation from extreme to moderate degree in one of their 
extreme tails - often refers to having 'asymmetric tail' - but with a 
fixed Kendall's $\tau$. One can control the degree of "tail-asymmetry" - moderate 
to extreme - using "copula" (for details see @Ghosh_copula and \nameref{Methods}), a statistical 
tool generally used to model dependencies between correlated variables. 
This 'asymmetric tail' is commonly found in many 
bivariate or multivariate dependence structures for environmental, species-trait, 
phenological, population, community, and ecosystem functioning datasets 
and can help to interpret the causal mechanisms as well as 
to draw consequential inferences for such patterns [@Ghosh_copula]. For example, it 
is well known that spatial dependence of environmental
variables is an important driver of spatial dependence in fluctuating 
population time series (called 'spatial synchrony'). Therefore, one may expect 
the synchronous populations over a larger network would suppress for extreme bad 
years of limiting resources than moderately 'good' years. Any two such populations, that are 
affected badly in a similar manner due to common food shortage, are expected to be more
"left-tail" associated (Figure \ref{fig_intro_cop}A, B) rather than "right-tail" (Figure \ref{fig_intro_cop}C, D) as the environmental fluctuation 
time-series have more "left-tail" association for those extreme bad years
(i.e., 'Moran-effect' like mechanism). Similarly, during 
favorable years if pest metapopulations 
grow abundantly compared to other average years, population time series look like 
\ref{fig_intro_cop}C or D with stronger "right-tail" association, turning local 
outbreaks into regional epidemics. In a 
nutshell, "tail-asymmetry" could have profound effect in understanding 
risk for population viability or ecosystem function
considering their relation with extreme climatic events.

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] used the above mentioned feature of "tail-asymmetry" 
to estimate extinction risk for a density-independent metapopulation model. The study 
showed "left-tail" associated density-independent metapopulations 
to be more prone to extinctions than "right-tail" associated one as we may expect species 
are more vulnerable to "catastrophs" (extreme bad years) than to "bonanzas" (extreme good years).
Present manuscript is an extension of that concept in estimating the extinction risk 
beyond the density-independent model. Population growth rate, when depending 
on its own density, serves as a more realistic model to regulate unbounded growth, hence 
models with "density-dependence" are often studied in population ecology to 
predict population viability over long-term period [@hassell1975density; @fowler1987review; 
@courchamp1999inverse]. Here, we will explore the general finding from @Ghosh_copula in the context of
the density-dependent metapopulation patches which experience same spatially correlated noise but having a variety of tail asymmetry (from extreme to moderate, Fig. \ref{fig_intro_cop}). It is expected that extreme left-tail associated noise might wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}).

<!--and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. Following we will reintroduce the concept of copula in a manner that follows @Ghosh_copula.-->

<!--***DAN: One thing you probably want to include is the logic (see BIVAN) that if extreme events are getting stronger and more extreme/intense events are also more spatially widespread, then it seems reasonable to expect that spatial patterns of tail associations in environmental variables will also be changing. You may also want to mention (one sentence only) that we use the statistical concept of the copula to control tail dependence in our simulations – more will be said about copulas in the Methods.-->

@@

From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail association between vital rates when temperature is too cold than normal) and 
thereby inflate the population viability.

Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}).

Though tail-association 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail association have never been considered 
in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.
